Fast Sorting on Modern Computers

### 2005-08-26 docs/tech_report/method.tex
minor edits due to david, and amortizZZZZZ

It takes time to fill arrays, and this can distort the relationship between the
results for small set sizes, and larger set sizes, for which this time is
amortized. As a result, we measured this time, and subtracted it from our
results.

>>>>

It takes time to fill arrays, and this can distort the relationship between the
results for small set sizes, and larger set sizes, for which this time is
amortised. As a result, we measured this time, and subtracted it from our
results.

### 2005-08-26 docs/tech_report/radix.tex
removed the word project, and resolved the occasional TODO
minor edits due to david, and amortizZZZZZ

All other sorts considered in this project are \n{comparison-based} sorts.
They compare two keys at a time, and determine the larger and smaller key,
sorting on that basis. Radixsort is a \n{counting-based} sort. It works by
separating the keys into small parts, and processing these parts without
regard for other keys. For radix of two, these small parts are bits, and for a
radix of 10, these parts would be digits.

@@

Memory-tuned radixsort should traverse the array only nine times: Once at the
start to count the least significant bytes; six times (once per array per pass)
as it copies using the three least significant bytes (counting the next least
significant byte at the same time), and a final two as it copies using the most
significant byte.

\subsection{Aligned Memory-tuned Radixsort}
In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
with reference to it auxiliary array. Specifically, extra memory is allocated
for the auxiliary array, which is then offset so that the start of the two
arrays are exactly half the length of the cache apart. We further added an
alignment for the level 1 cache (see Section \ref{double-aligned tiled
mergesort}) to significantly reduce level 1 misses at the expense of a small
number of level 2 misses. We adapt this strategy here, and call it Aligned
Memory-tuned Radixsort.

@@

The number of branch prediction misses should be exceptionally low. In
comparison-based sorts, a branch due to a comparison can not be predicted except
in exceptional circumstances\footnote{See Section \ref{insertion is
predictable}.}. Radixsort has no comparative branches at all. As a result, only
flow control branches occur in radixsort, and they are all predictable.

@@

The number of branch misses, shown in Figure \ref{radix branch misses} was also
extremely low. Despite three branches per key, there was less than .02 of a branch
miss per key. The number of branches itself is quite low, as several of the
sorts considered had nearly that number of misses. However, it is not
unexpected. The number of branches reduces when copying back is removed, and
while this reduces the number of mispredictions, there are so few to begin with
that this result is not significant.

@@

LaMarca's radixsort results are quite different to those presented here. His
instruction count graph has a logarithmic shape: it begins very high and sharply
falls before flattening out. By contrast, the graph here is almost completely
flat. He shows a similar result in both cache miss graphs: a sharp dip, followed
by a nearly flat-line. This may be due to amortization, in which case the trend
would not be expected in our graphs due to subtraction of the time taken to fill
the array.

>>>>

All other sorts considered in this report are \n{comparison-based} sorts.
They compare two keys at a time, and determine the larger and smaller key,
sorting on that basis. Radixsort is a \n{counting-based} sort. It works by
separating the keys into small parts, and processing these parts without
regard for other keys. For radix of two, these small parts are bits, and for a
radix of 10, these parts would be digits.

@@

For unsigned 32-bit integers, memory-tuned radixsort should traverse the two
array only nine times in total: Once at the start to count the least significant
bytes; six times (once per array per pass) as it copies using the three least
significant bytes (counting the next least significant byte at the same time),
and a final two as it copies using the most significant byte.

\subsection{Aligned Memory-tuned Radixsort}
In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
with reference to its auxiliary array. Specifically, extra memory is allocated
for the auxiliary array, which is then offset so that the start of the two
arrays are exactly half the length of the cache apart. We further added an
alignment for the level 1 cache (see Section \ref{double-aligned tiled
mergesort}) to significantly reduce level 1 misses at the expense of a small
number of level 2 misses. We adapt this strategy here, and call it Aligned
Memory-tuned Radixsort.

@@

The number of branch prediction misses should be exceptionally low. In
comparison-based sorts, there are generally frequent mispredictions\footnote{See
Section \ref{insertion is predictable} for an exception.}. Radixsort has no
comparative branches at all. As a result, only flow control branches occur in
radixsort, and they are all predictable.

@@

The number of branch misses, shown in Figure \ref{radix branch misses} was also
extremely low. Despite three branches per key, there was less than .02 of a branch
miss per key. The number of branches itself is quite low, as several of the
sorts considered had more thna that number of mispredictions. However, it is not
unexpected. The number of branches reduces when copying back is removed, and
while this reduces the number of mispredictions, there are so few to begin with
that this result is not significant.

@@

LaMarca's radixsort results are quite different to those presented here. His
instruction count graph has a logarithmic shape: it begins very high and sharply
falls before flattening out. By contrast, the graph here is almost completely
flat. He shows a similar result in both cache miss graphs: a sharp dip, followed
by a nearly flat-line. This is due to the instructions, time and compulsory
misses being amortised over the ever larger array size. The trend is not
expected in our graphs due to subtraction of the time taken to fill the array.
